MADA Project Part 2

### 2024-04-21 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 5
Analysis is complete. Readme files and removal of outdated datasets and codes will be removed throughout the week. Additionally, the manuscript will be finalized this weel

## This file contains the process for loading, cleaning and exploring the datasets.

@@

# Correctly expand "Both" in Test Type, then adjust to "EEE" and "WNV"
pcr_data_expanded <- pcr_data %>%
  mutate(Test_Type_Expanded = case_when(
    `Test Type` == "Both" ~ "EEE&WNV", # Temporarily mark "Both" for special handling
    TRUE ~ `Test Type`
  )) %>%
  separate_rows(Test_Type_Expanded, sep = "&") %>%
  mutate(
    Year = year(`Collection Date`),
    Week = week(`Collection Date`),
    Submitted_for_Testing = `Submitted for Testing` == "Y",
    # Assuming positive results are marked as "Positive" in `Result`
    Is_Positive = Result == "Positive"
  )

# Save as RDS
saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data_expanded.rds"))

# Save as CSV
write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data_expanded.csv"))

@@

### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing

### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data

### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020

### This code will load all the files, combine them, and then save the combined file in the "raw data" folder

```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

@@

# Ensure hourly_wx is read if not already done
hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)

hourly_wx_clean <- hourly_wx %>%
  mutate(Year = year(DATE),
         ISO_Week = isoweek(DATE),
         week_num = sprintf("%d-%02d", Year, ISO_Week)) %>%
  filter(REPORT_TYPE...3 != "SOD", REPORT_TYPE...3 != "SOM", REPORT_TYPE...3 != "AUTO") %>%
  select(STATION, DATE, HourlyDryBulbTemperature, HourlyRelativeHumidity, HourlyWetBulbTemperature, HourlyWindSpeed, week_num)

@@

```
## Drought data
### These datasets were obtained from the US Drought Monitor
#### The U.S. Drought Monitor is produced through a partnership between the National Drought Mitigation Center at the University of Nebraska-Lincoln, the United States Department of Agriculture and the National Oceanic and Atmospheric Administration
##### This data is organized by geographic area and week
##### It includes 4 categories. D4 - Exceptional Drought (0-2 percentile), D3 - Extreme Drought (2-5 percentile), D2 - Severe Drought (5-10 percentile), D1 - Moderate Drought (10-20 percentile), D0 - Abnormally Dry (20-30 percentile), None or 0 - Neither Wet or Dry
##### It provides the percent area covered by each category. 
```{r}
# read file with drought data by county for MA, RI, CT for the years 2007-2023
drought_ma_ri_ct = read_csv(here("data", "raw-data", "comp_stat_county_ma_ri_ct.csv"), show_col_types = FALSE)

# read file with drought data by county for FEMA Region 4 for the years 2007-2023
drought_fema4 = read_csv(here("data", "raw-data", "comp_stat_fema4.csv"), show_col_types = FALSE)

# Add week number based on MapDate for both files
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
drought_ma_ri_ct <- drought_ma_ri_ct %>%
  mutate(
    # First, ensure MapDate is in a proper date format
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Then, format the collection date into "YYYY-WW"
    week_num = paste0(year(MapDate), "-", 
                      sprintf("%02d", isoweek(MapDate)))
  )

drought_fema4 <- drought_fema4 %>%
  mutate(
    # First, ensure MapDate is in a proper date format
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Then, format the collection date into "YYYY-WW"
    week_num = paste0(year(MapDate), "-", 
                      sprintf("%02d", isoweek(MapDate)))
  )

# Define the path for the RDS and CSV files
rds_path_drought_ma_ri_ct<- here("data", "processed-data", "rds", "drought_ma_ri_ct.rds")
csv_path_drought_ma_ri_ct <- here("data", "processed-data", "drought_ma_ri_ct.csv")
# Save the wx_selected dataframe as an RDS and CSV files
saveRDS(drought_ma_ri_ct, rds_path_drought_ma_ri_ct)
write.csv(drought_ma_ri_ct, csv_path_drought_ma_ri_ct, row.names = FALSE)

# Define the path for the RDS and CSV files
rds_path_drought_fema4<- here("data", "processed-data", "rds", "drought_fema4.rds")
csv_path_drought_fema4 <- here("data", "processed-data", "drought_fema4.csv")
# Save the wx_selected dataframe as an RDS and CSV files
saveRDS(drought_fema4, rds_path_drought_fema4)
write.csv(drought_fema4, csv_path_drought_fema4, row.names = FALSE)
```

@@

# Processing and cleaning the Mosquito trap data

### Examine the variables and change all column names to lower case

```{r}
str(mosquito_raw)

@@

## we will examine the dataset before making changes.

### first we will Identify missing data

```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

@@

# Ensure there's no duplication in the conversion attempt
# Step 3: Convert list columns to numeric, replacing NULL or list elements without values with 0
# Utilizing sapply for a direct approach to handle potential list-columns
wide_data <- wide_data %>%
  mutate(across(.cols = all_of(species_columns), 
                .fns = ~ map_dbl(.x, function(item) {
                  # Handle NULL items directly
                  if (is.null(item)) {
                    return(0)
                  # Handle vectors: convert NA to 0, and take the first item if it's a vector
                  } else if (length(item) > 1 || is.na(item)) {
                    item[is.na(item)] <- 0
                    return(as.numeric(item[1]))
                  } else {
                    return(as.numeric(item))
                  }
                })))
# Step 4: Add week_num column
wide_data <- wide_data %>%
  mutate(
    `collection date` = as.Date(`collection date`),
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )

# Output the transformed data to inspect its structure
head(wide_data)

@@

# Now we will filter the the mosquito data to match the time period of the virus data. 
# Since mosquito populations can be affected by the previous year's populations, we will filter for 2011-

@@

# Save the filtered wide-format data to the specified CSV path
write_csv(filtered_mosquito_data, file_path_filtered_csv)

# Define the path for the RDS file
file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")

@@

# Define path to the data file
trap_wide_path <- here("data", "processed-data", "mosquito_wide.csv")

@@

# Calculate the proportion of MEL out of the total for all species
proportion_mel_by_year <- trap_proportion %>%
  group_by(Year) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(MEL + PER + PIP + RES + SAL + TRI + SAP, na.rm = TRUE),
            MEL_Prop = Total_MEL / Total_All)

@@

### This will load sunset and sunrise times.
#### the amount of sunlight affects mosquitoes activities and provides cues for them to enter diapause.
##### The suncalc package will be used to calculate the sunrise and sunset times for Bristol County, MA, which is the study location. The times will be calculated for each day from 2007 to 2023. The data will be saved as a CSV file and an RDS file for future use.
```{r}
# Ensure the suncalc library is loaded
library(suncalc)
get_sun_times <- function(date, latitude, longitude) {
  # Get the sunrise and sunset times from the suncalc package
  times <- getSunlightTimes(date = date, lat = latitude, lon = longitude, tz = "UTC")

@@

#Average daylight by week_num
sun_times_df <- sun_times_df %>%
  mutate(
    # Combine Date with Sunrise and Sunset times to create full POSIXct datetime objects
    Sunrise_dt = as.POSIXct(paste(Date, Sunrise), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
    Sunset_dt = as.POSIXct(paste(Date, Sunset), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
    # Now calculate the difference in hours between Sunset and Sunrise
    daylight_hours = as.numeric(difftime(Sunset_dt, Sunrise_dt, units = "hours"))
  )

# Aggregate data by week and calculate average daylight hours
average_daylight_by_week <- sun_times_df %>%
  group_by(week_num) %>%
  summarise(average_daylight_hours = mean(daylight_hours))

# Print first few rows of the resulting data frame
head(average_daylight_by_week)

>>>>

## This file contains the process for loading, cleaning and exploring the datasets.
### There is also a little visualizing data

@@

# Save as RDS
saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data"))

# Save as CSV
write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data.csv"))

@@

### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing
### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

@@

# Ensure hourly_wx is read if not already done
hourly_wx <- read_csv(hourly_wx_path, show_col_types = FALSE)
View(hourly_wx)
hourly_wx_clean <- hourly_wx %>%
  mutate(Year = year(DATE),
         ISO_Week = isoweek(DATE),
         week_num = sprintf("%d-%02d", Year, ISO_Week)) %>%
  filter(REPORT_TYPE...3 != "SOD", REPORT_TYPE...3 != "SOM", REPORT_TYPE...3 != "AUTO") %>%
  select(STATION, DATE, HourlyDryBulbTemperature, HourlyRelativeHumidity, HourlyWetBulbTemperature, HourlyWindSpeed, week_num)

@@

```
## Drought data
### These datasets were obtained from the US Drought Monitor
#### The U.S. Drought Monitor is produced through a partnership between the National Drought Mitigation Center at the University of Nebraska-Lincoln, the United States Department of Agriculture and the National Oceanic and Atmospheric Administration
##### This data is organized by geographic area and week
##### It includes 4 categories. D4 - Exceptional Drought (0-2 percentile), D3 - Extreme Drought (2-5 percentile), D2 - Severe Drought (5-10 percentile), D1 - Moderate Drought (10-20 percentile), D0 - Abnormally Dry (20-30 percentile), None or 0 - Neither Wet or Dry
##### It provides the percent area covered by each category. 
```{r}
# read file with cumulative drought data by county for all MA counties for the years 2006-2023
drought_ma_counties = read_csv(here("data", "raw-data", "comp_stat_ma_counties.csv"), show_col_types = FALSE)

# read file with drought data by county for FEMA Region 1-4 for the years 2007-2023
drought_fema1_4 = read_csv(here("data", "raw-data", "comp_stat_fema1_4.csv"), show_col_types = FALSE)
#View(drought_fema1_4)
# Add week number based on MapDate for both files
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08

# Additionally, this code calculates the Drought Severity and Coverage Index (DSCI) for each dataset
# The DSCI is calculated as the sum of D0, D1, D2, D3, and D4 categories
# This index provides a measure of the overall drought severity and can range from 0-500,
# where 0 is no drought and 500 is the most severe drought over the entire area

drought_ma_counties <- drought_ma_counties %>%
  mutate(
    # Convert MapDate from a numeric format to a Date object
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Create a week_num column formatted as "YYYY-WW"
    week_num = paste0(year(MapDate), "-", sprintf("%02d", isoweek(MapDate))),
    # Calculate the DSCI, assuming D0 to D4 are numeric
    DSCI_ma = rowSums(select(., D0:D4), na.rm = TRUE)
  )
#View(drought_ma_counties)

drought_fema1_4 <- drought_fema1_4 %>%
 mutate(
    # Convert MapDate from a numeric format to a Date object
    MapDate = as.Date(as.character(MapDate), format = "%Y%m%d"),
    # Create a week_num column formatted as "YYYY-WW"
    week_num = paste0(year(MapDate), "-", sprintf("%02d", isoweek(MapDate))),
    # Calculate the DSCI, assuming D0 to D4 are numeric
        DSCI_fema = rowSums(select(., c("D0", "D1", "D2", "D3", "D4")))
  )

# Define the path for the RDS and CSV files
rds_path_drought_ma_counties<- here("data", "processed-data", "rds", "drought_ma_counties.rds")
csv_path_drought_ma_counties <- here("data", "processed-data", "drought_ma_counties.csv")
# Save the dataframe as an RDS and CSV files
saveRDS(drought_ma_counties, rds_path_drought_ma_counties)
write.csv(drought_ma_counties, csv_path_drought_ma_counties, row.names = FALSE)

# Define the path for the RDS and CSV files
rds_path_drought_fema1_4<- here("data", "processed-data", "rds", "drought_fema1_4.rds")
csv_path_drought_fema1_4 <- here("data", "processed-data", "drought_fema1_4.csv")
# Save the wx_selected dataframe as an RDS and CSV files
saveRDS(drought_fema1_4, rds_path_drought_fema1_4)
write.csv(drought_fema1_4, csv_path_drought_fema1_4, row.names = FALSE)
```

@@

## Bird data
#### This dataset comes from the annual North American Breeding Bird Survey that is conducted in a joint effort by: USGS Eastern Ecological Science Center, Environment and Climate Change Canada, and the Mexican National Commission for the Knowledge and Use of Biodiversity.
#### There are 2 (5 and 11) routes in Bristol County, MA. The data is from 2007-2022. However, route 11 is mission data for 2013 and route 5 ends in 2019.
#### For route 11, 2013, we will use the data from 2012 and 2014 to estimate the missing data.
for route 5, 2020-2022, we will use the data from 2019 to estimate the missing data.
```{r}
# Load the dataset
file_path_bird <- here("data", "raw-data", "Massach.csv")
bird_data_raw <- read.csv(file_path_bird)

# Filter and clean up the data
bird_bbs <- bird_data_raw %>%
  filter(Route %in% c(5, 11), Year >= 2006 & Year <= 2022) %>%
  select(-AOU, -Count10, -Count20, -Count30, -Count40, -Count50)

# Summarize the data
bird_bbs_summarized <- bird_bbs %>%
  group_by(Route, Year) %>%
  summarize(
    StopTotal = sum(StopTotal, na.rm = TRUE),
    SpeciesTotal = sum(SpeciesTotal, na.rm = TRUE),
    .groups = 'drop'
  )

# Check for missing data within Routes 5 and 11
missing_data <- map_df(c(5, 11), ~tibble(
  Route = .x,
  missing_years = setdiff(2006:2022, bird_bbs_summarized$Year[bird_bbs_summarized$Route == .x])
))

# Identify all combinations of Route and Year within the range
all_combinations <- expand.grid(Route = unique(bird_bbs_summarized$Route),
                                Year = 2006:2022)

# Join with the summarized data to identify missing rows
bird_bbs_full <- left_join(all_combinations, bird_bbs_summarized, by = c("Route", "Year"))

# Impute missing data by taking the average of the previous and next year for each route
bird_bbs_imputed <- bird_bbs_full %>%
  group_by(Route) %>%
  arrange(Route, Year) %>%
  mutate(
    StopTotal = ifelse(is.na(StopTotal), (lag(StopTotal) + lead(StopTotal)) / 2, StopTotal),
    SpeciesTotal = ifelse(is.na(SpeciesTotal), (lag(SpeciesTotal) + lead(SpeciesTotal)) / 2, SpeciesTotal)
  ) %>%
  ungroup()

# Handling the last records for Route 5 (2020-2022)
# Since there's no future data to average, you might replicate the most recent data (e.g., from 2019)
# Or consider another imputation method suitable for your analysis and data characteristics

# Find the most recent year's data for Route 5
recent_data_route_5 <- bird_bbs_imputed %>% 
  filter(Route == 5, Year == max(Year[Route == 5 & !is.na(StopTotal)]))

# Create a template for the missing years
missing_years_route_5 <- tibble(Route = 5, Year = 2020:2022)

# Replicate the most recent data for the missing years
imputed_missing_years_route_5 <- bind_rows(lapply(missing_years_route_5$Year, function(y) {
  mutate(recent_data_route_5, Year = y)
})) %>%
  bind_rows(bird_bbs_imputed) %>%
  arrange(Route, Year)

# Check if any adjustments or specific imputations are needed
print(imputed_missing_years_route_5)

bird_clean <- imputed_missing_years_route_5 %>%
  drop_na()

# Save the cleaned bird data as an csv
write_csv(bird_clean, here("data", "processed-data", "bird_clean.csv"))
```

# Processing and cleaning the Mosquito trap data
### Examine the variables and change all column names to lower case
```{r}
str(mosquito_raw)

@@

## we will examine the dataset before making changes.
### first we will Identify missing data
```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

@@

# Ensure there's no duplication in the conversion attempt
# Step 3: Convert list columns to numeric, replacing NULL or list elements without values with 0
# Utilizing sapply for a direct approach to handle potential list-columns
wide_data <- wide_data %>%
  mutate(across(.cols = all_of(species_columns), 
                .fns = ~ map_dbl(.x, function(item) {
                  # Handle NULL items directly
                  if (is.null(item)) {
                    return(0)
                  # Handle vectors: convert NA to 0, and take the first item if it's a vector
                  } else if (length(item) > 1 || is.na(item)) {
                    item[is.na(item)] <- 0
                    return(as.numeric(item[1]))
                  } else {
                    return(as.numeric(item))
                  }
                })))
# Step 4: Add week_num column
mosquito_wide <- wide_data %>%
  mutate(
    `collection date` = as.Date(`collection date`),
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )

# Output the transformed data to inspect its structure
head(mosquito_wide)

@@

# Now we will filter the the mosquito data to match the time period of the virus data. 
# Since mosquito populations can be affected by the previous year's populations, we will filter for 2007-2023

@@

# Save the filtered wide-format data to the specified CSV path
write_csv(filtered_mosquito_data, file_path_filtered_csv)
View(mosquito_wide_filtered.csv)
# Define the path for the RDS file
file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")

@@

# Define path to the data file
trap_wide_path <- here("data", "processed-data", "mosquito_wide_all.csv")

@@

# Assuming all species columns are next to each other, calculate the proportion of MEL
proportion_mel_by_year <- trap_proportion %>%
  # Calculate row-wise totals for all species
  rowwise() %>%
  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
  ungroup() %>%
  # Then group by year and summarize
  group_by(Year) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by year
            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL

# Print the results
print(proportion_mel_by_year)

# Calculate the proportion of MEL by week
# Assuming all species columns are next to each other, calculate the proportion of MEL by week_num
proportion_mel_by_week <- trap_proportion %>%
  # Calculate row-wise totals for all species
  rowwise() %>%
  mutate(Total_All = sum(c_across(ABS:OCH), na.rm = TRUE)) %>%
  ungroup() %>%
  # Then group by week_num and summarize
  group_by(week_num) %>%
  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
            Total_All = sum(Total_All, na.rm = TRUE), # Sum of all species by week_num
            MEL_Prop = Total_MEL / Total_All) # Proportion of MEL

# Print the results
print(proportion_mel_by_week)

# save the data as CSV
write_csv(proportion_mel_by_week, here("data", "processed-data", "mel_proportion_by_week.csv"))

@@

### This will load sunset and sunrise times from the suncalc package. However, it can be slow to complete.
#### the amount of sunlight affects mosquitoes activities and provides cues for them to enter diapause.
##### The suncalc package will be used to calculate the sunrise and sunset times for Bristol County, MA, which is the study location. The times will be calculated for each day from 2007 to 2023. The data will be saved as a CSV file and an RDS file for future use.

```{r}
# Ensure the suncalc library is loaded
library(suncalc)
get_sun_times <- function(date, latitude, longitude) {
  # Get the sunrise and sunset times from the suncalc package
  times <- getSunlightTimes(date = date, lat = latitude, lon = longitude, tz = "UTC")

@@

#here
# Add 'week_num' calculation here
sun_times_df <- sun_times_df %>%
  mutate(
    # Convert Date column to Date object if it's not already
    Date = as.Date(Date),
    Sunrise_dt = as.POSIXct(paste(Date, Sunrise), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
    Sunset_dt = as.POSIXct(paste(Date, Sunset), format="%Y-%m-%d %H:%M:%S", tz="UTC"),
    # Calculate daylight hours
    daylight_hours = as.numeric(difftime(Sunset_dt, Sunrise_dt, units = "hours")),
    # Calculate week number as "YYYY-WW"
    week_num = paste(year(Date), sprintf("%02d", isoweek(Date)), sep = "-")
  )

# Now, proceed to aggregate by 'week_num'
average_daylight_by_week <- sun_times_df %>%
  group_by(week_num) %>%
  summarise(average_daylight_hours = mean(daylight_hours, na.rm = TRUE))  # Ensure NA values are handled

# Optionally, save or print your aggregated data
write.csv(average_daylight_by_week, here("data", "processed-data", "average_daylight_by_week.csv"), row.names = FALSE)
# Print the first few rows to verify
head(average_daylight_by_week)
